---
title: "scPDSI tests"
author: "Carlos Dobler"
date: "1/31/2022"
output: github_document
---

Here I show results of a series of test runs for calculating scPDSI with the R package `scpdsi`. All tests were ran for a region centered on Mexico to reduce processing times. Of the variables that scPDSI asks for, the only ones that would vary were monthly precipitation and monthly potential evapotranspiration (PET). AWC remained constant (100 mm), and the start and end dates for calibration included the whole period. I used four PET formulations: one uses the "raw" potential evaporation from ERA5 Reanalysis data; the other three were calculated with the R package `SPEI` using various ERA5 Reanalysis variables. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

source(here::here("scripts/00-set-up.R"))

library(SPEI)
library(scPDSI)
library(colorspace)

plan(multicore, workers = 6)

# mex
c(st_point(c(-120, 33)),
  st_point(c(-80, 13))) %>% 
  st_bbox() %>% 
  st_set_crs(4326) -> lim

seq(as_date("1979-01-01"), as_date("2020-12-01"), by = "months") -> date_vector

# READ ALL VARS

# precip
readRDS(here::here("output", "starr_precip_monthlysums_monthlyres.rds")) %>% 
  st_crop(lim) -> var_pr

# potential evaporation
readRDS(here::here("output", "starr_potevap_monthlysums_monthlyres.rds")) %>% 
  st_crop(lim)-> var_potev

# external radiation
read_ncdf(str_glue("{dir_bucket_mine}/era/monthly/era5_monthly_mean_daily_radiation.nc")) %>% 
  st_crop(lim) -> var_r

var_r %>%   
  select(tisr) %>% # top (external?)
  mutate(tisr = units::set_units(tisr, MJ/m^2),
         tisr = units::drop_units(tisr)) -> var_ra

# internal radiation
var_r %>% 
  select(ssrd) %>%
  mutate(ssrd = units::set_units(ssrd, MJ/m^2),
         ssrd = units::drop_units(ssrd)) -> var_rs

rm(var_r)

# average tmperature
read_ncdf(str_glue("{dir_bucket_mine}/era/monthly/era5_2mtemp_dailymeans_monthlyres_rawfile.nc")) %>% 
  st_crop(lim) %>% 
  mutate(t2m = units::set_units(t2m, degC),
         t2m = units::drop_units(t2m)) -> var_tave

# latitude
var_tave %>% 
  slice(time,1) %>%
  st_dim_to_attr(2) -> var_lat

# maximum temperature
map(seq_len(42), function(y){
  
  list.files("~/bucket_mine/era/monthly/2m_temperature_max/", full.names = T) %>% 
    .[y] %>% 
    readRDS() %>% 
    st_crop(lim)
  
})  %>% 
  do.call(c, .) -> var_tmax

# minimum temperature
map(seq_len(42), function(y){
  
  list.files("~/bucket_mine/era/monthly/2m_temperature_min/", full.names = T) %>% 
    .[y] %>% 
    readRDS() %>% 
    st_crop(lim)
  
}) %>% 
  do.call(c, .) -> var_tmin

# wind speed
# need to convert to 2m
read_ncdf(str_glue("{dir_bucket_mine}/era/monthly/era5_monthly_mean_10mwindspeed.nc")) %>% 
  st_crop(lim) %>% 
  mutate(si10 = units::drop_units(si10)) -> var_u10

# dewpoint temp
read_ncdf(str_glue("{dir_bucket_mine}/era/monthly/era5_monthly_mean_dewpointtemp.nc")) %>% 
  st_crop(lim) %>% 
  mutate(d2m = units::set_units(d2m, degC),
         d2m = units::drop_units(d2m)) -> var_dewptemp

# surface pressure
read_ncdf(str_glue("{dir_bucket_mine}/era/monthly/era5_monthly_mean_surfpressure.nc")) %>% 
  st_crop(lim) -> var_pres

var_pres %>% 
  mutate(sp = units::set_units(sp, kPa),
         sp = units::drop_units(sp)) -> var_pres

# ***

# van der schrier
read_ncdf(here::here("data", "scPDSI.cru_ts4.05early1.1901.2020.cal_1901_20.bams.2021.GLOBAL.1901.2020.nc")) %>%
  st_warp(var_pr) %>% 
  st_crop(lim) %>% 
  filter(year(time) >= 1979) %>% 
  st_set_dimensions("time", values = date_vector) -> vds


```



## 1. ...with ERA5's potential evaporation

```{r}

c(var_pr, var_potev, along = 3) %>% 
  st_apply(c(1,2), function(x){
    
    pdsi(P = x[1:504], 
         PE = x[505:1008], 
         sc = T)$X %>% as.vector()
    
  },
  FUTURE = T,
  future.seed = NULL,
  .fname = "time") %>% 
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> s_pdsi

s_pdsi[is.na(adrop(vds[,,,1]))] <- NA

```

Mapping a random date:
```{r}

sample(date_vector, 1) -> d

ggplot() +
  geom_stars(data = s_pdsi %>% 
               filter(time == d) %>% 
               adrop() %>% 
               mutate(X = case_when(X < -5 ~ -5,
                                    X > 5 ~ 5,
                                    TRUE ~ X))) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80",
                                  name = "PDSI"
                                  ) +
  coord_fixed() +
  labs(subtitle = d)

```

### 1.1. Temporal correlation
The following figure correlates my results against Van der Schrier's PDSI time series on a per-pixel basis. Overall, correlation values are low, and in some regions (e.g. Arizona) are negative. 

```{r}

func_t_cor_map <- function(era_map, vds_map){
  
  c(era_map, vds_map, along = 3) %>% 
  st_apply(c(1,2), function(x){
    
    cor(x[1:504], x[505:1008])
    
  }, 
  FUTURE = T)
}

func_t_cor_map(s_pdsi, vds) -> cor_map

ggplot() +
  geom_stars(data = cor_map) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80", 
                                  limits = c(-0.95,0.95),
                                  name = "r") +
  coord_fixed()

```

If I randomly choose a pixel with a HIGH correlation coefficient, the time series look like this:

```{r}
func_ts_comparison <- function(thres){
  
  cor_map %>% 
    as_tibble() %>% 
    filter(near(X, thres, 0.01)) %>% 
    slice_sample(n = 1) %>% 
    {c(.$longitude, .$latitude)} -> coords
  
  vds %>%
    as_tibble() %>% 
    filter(longitude == coords[1],
           latitude == coords[2]) %>% 
    pull(scpdsi) -> vect_vds

  s_pdsi %>% 
    as_tibble() %>% 
    filter(longitude == coords[1],
           latitude == coords[2]) %>% 
    pull(X) -> vect_era

  tibble(vds = vect_vds,
         era5 = vect_era,
         time = date_vector) %>% 
    pivot_longer(-time, names_to = "src", values_to = "pdsi") %>% 
    
    ggplot(aes(x = time, y = pdsi, color = src)) +
    geom_line()
  
}

func_ts_comparison(0.8) 

```

If I choose one with a LOW correlation coefficient:
```{r}
func_ts_comparison(0.1) 
```

If I choose one with a NEGATIVE correlation coefficient:
```{r}
func_ts_comparison(-0.3)
```

# 1.2. Spatial correlation
The following figure shows correlation coefficients between my results and Van der Schrier's PDSI at a month level (i.e. my resulting map at t-n vs. VDS resulting map at t-n). We can see that correlation fluctuates over time, between a coefficient of ~0 and ~0.5.
```{r}

c(s_pdsi, vds, along = list(foo = c("vds", "era5"))) %>% 
  split("foo") %>% 
  as_tibble() %>% 
  group_by(time) %>% 
  summarize(cor = cor(era5, vds, use = "complete.obs")) -> t_cor 

t_cor %>%   
  ggplot(aes(x = time, y = cor)) +
  geom_point() +
  geom_smooth()

```

If I randomly choose a date where correlation was HIGH-ish, the two maps look like this:
```{r}
func_sp_comparison <- function(thres){
  
  t_cor %>% 
    as_tibble() %>% 
    filter(near(cor, thres, 0.01)) %>% 
    slice_sample(n = 1) %>% 
    pull(time) -> t_time
  
  s_pdsi %>%
    filter(time == t_time) %>% 
    adrop() %>% 
    as_tibble() -> m_1
  
  vds %>%
    filter(time == t_time) %>% 
    adrop() %>% 
    as_tibble() -> m_2
  
  left_join(m_1, m_2, by = c("longitude", "latitude")) %>%
    rename(era5 = 3,
           vds = 4) %>% 
    pivot_longer(3:4, names_to = "src", values_to = "pdsi") %>% 
    mutate(pdsi = case_when(pdsi < -5 ~ -5,
                            pdsi > 5 ~ 5,
                            TRUE ~ pdsi)) %>% 
    ggplot(aes(x = longitude, y = latitude, fill = pdsi)) +
    geom_raster() +
    scale_fill_continuous_diverging(palette = "Blue-Red", rev = T, na.value = "grey80") +
    facet_wrap(~src, ncol = 2) +
    coord_equal() +
    labs(subtitle = t_time)
  
}

func_sp_comparison(0.5)

```

If I choose one where correlation is LOW:
```{r}
func_sp_comparison(0.1)
```




## 2. ...with Thornthwaite
Thornthwaite's PET formulation uses two variables: average temperature and latitude.
```{r}

# calculate pet
var_tave %>% # average temp
  split("time") %>% 
  c(var_lat) %>% # lat
  merge() %>% 
  
  st_apply(c(1,2), function(x){
    
    x[length(x)] -> lat
    x[-length(x)] -> x_ts
    
    thornthwaite(Tave = x_ts, 
                 lat = lat)
    
  }, 
  FUTURE = T,
  .fname = "time") %>% 
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> pet

# calculate pdsi
c(var_pr, pet, along = 3) %>% 
  st_apply(c(1,2), function(x){
    
    pdsi(P = x[1:504], 
         PE = x[505:1008], 
         sc = T)$X %>% as.vector()
    
  },
  FUTURE = T,
  future.seed = NULL,
  .fname = "time") %>% 
  
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> s_pdsi

```

Mapping a random date:
```{r}

s_pdsi[is.na(adrop(vds[,,,1]))] <- NA

sample(date_vector, 1) -> d

ggplot() +
  geom_stars(data = s_pdsi %>% 
               filter(time == d) %>% 
               adrop() %>% 
               mutate(X = case_when(X < -5 ~ -5,
                                    X > 5 ~ 5,
                                    TRUE ~ X))) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80",
                                  name = "PDSI") +
  coord_fixed() +
  labs(subtitle = d)

```


### 2.1. Temporal correlation
When correlating my results using Thornthwaite vs Van der Schrier's, we see an improvement over previous results. Almost all cells show positive correlations, although they tend to be quite low in the south-east of Mexico.
```{r}

func_t_cor_map(s_pdsi, vds) -> cor_map

ggplot() +
  geom_stars(data = cor_map) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80", 
                                  limits = c(-0.95,0.95),
                                  name = "r") +
  coord_fixed()

```

If I randomly choose a pixel with HIGH correlation, the time series look like this:
```{r}
func_ts_comparison(0.8) 
```

And with LOW correlation:
```{r}
func_ts_comparison(0.1)
```

# 2.2. Spatial correlation
Spatially, correlation between my results and Van der Schrier's fluctuates around ~0.1 and ~0.6:
```{r}

c(s_pdsi, vds, along = list(foo = c("vds", "era5"))) %>% 
  split("foo") %>% 
  as_tibble() %>% 
  group_by(time) %>% 
  summarize(cor = cor(era5, vds, use = "complete.obs")) -> t_cor 

t_cor %>%   
  ggplot(aes(x = time, y = cor)) +
  geom_point() +
  geom_smooth()

```

If I randomly choose a date where correlation was HIGH:
```{r}
func_sp_comparison(0.6)
```
And a date where correlation was LOW:
```{r}
func_sp_comparison(0.1)
```



## 3. ...with Hargreaves
Hargraves' PET formulation uses three variables: maximum temperature, minimum temperature, and radiation.

```{r}

# pet
c(var_ra, var_tmax, var_tmin) %>%
  st_apply(c(1,2), function(x){
    
    hargreaves(Tmin = x[1009:1512],
               Tmax = x[505:1008],
               Ra = x[1:504],
               na.rm = T)
    
  },
  FUTURE = T,
  .fname = "time") %>% 
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> pet

# pdsi
c(var_pr, pet, along = 3) %>% 
  st_apply(c(1,2), function(x){
    
    pdsi(P = x[1:504], 
         PE = x[505:1008], 
         sc = T)$X %>% as.vector()
    
  },
  FUTURE = T,
  future.seed = NULL,
  .fname = "time") %>% 
  
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> s_pdsi

```

Mapping a random date:
```{r}
s_pdsi[is.na(adrop(vds[,,,1]))] <- NA

sample(date_vector, 1) -> d

ggplot() +
  geom_stars(data = s_pdsi %>% 
               filter(time == d) %>% 
               adrop() %>% 
               mutate(X = case_when(X < -5 ~ -5,
                                    X > 5 ~ 5,
                                    TRUE ~ X))) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80",
                                  name = "PDSI") +
  coord_fixed() +
  labs(subtitle = d)

```


### 3.1. Temporal correlation
Using Hargraves doesn't seem to improve much the correlation compared to those obtained using Thorthwaite. Most pixels show positive correlation, but again, values in the south-east tend to be low.
```{r}

func_t_cor_map(s_pdsi, vds) -> cor_map

ggplot() +
  geom_stars(data = cor_map) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80", 
                                  limits = c(-0.95,0.95),
                                  name = "r") +
  coord_fixed()

```

If I randomly choose a pixel with HIGH correlation:
```{r}
func_ts_comparison(0.8) 
```

And one with LOW correlation:
```{r}
func_ts_comparison(0.1) 
```

# 3.2. Spatial correlation
Interestingly, spatial correlation between my results and Van der Schrier's shows a downward trend over time. 
```{r}

c(s_pdsi, vds, along = list(foo = c("vds", "era5"))) %>% 
  split("foo") %>% 
  as_tibble() %>% 
  group_by(time) %>% 
  summarize(cor = cor(era5, vds, use = "complete.obs")) -> t_cor 

t_cor %>%   
  ggplot(aes(x = time, y = cor)) +
  geom_point() +
  geom_smooth()

```

If I randomly choose a date with a HIGH-ish correlation:
```{r}
func_sp_comparison(0.55)
```

And one with LOW correlation:
```{r}
func_sp_comparison(0.1)
```



## 4. ...with Penman
Penman-Monteith's PET formulation can include several variables; some are used to derive others. In this case, I used: maximum temperature, minimum temperature, wind speed (it should be at 2m but I used at 10m), external radiance (I used "top of atmosphere"), incoming radiance, dewpoint temperature (to derive vapor pressure), and surface pressure. The function recommends to include elevation, something I'll try later.  

```{r}

# pet
source(here::here("scripts", "penman_mod.R"))

c(var_tmin, var_tmax, var_u10, var_ra, var_rs, var_dewptemp, var_pres, along = 3) %>%
  st_apply(c(1,2), function(x){
    
    penman_mod(Tmin = x[1:504],
           Tmax = x[505:1008],
           U2 = x[1009:1512],
           Ra = x[1513:2016],
           Rs = x[2017:2520],
           Tdew = x[2521:3024],
           P = x[3025:3528])
    
  },
  FUTURE = T,
  .fname = "time") %>% 
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> pet

# pdsi
c(var_pr, pet, along = 3) %>% 
  st_apply(c(1,2), function(x){
    
    pdsi(P = x[1:504], 
         PE = x[505:1008], 
         sc = T)$X %>% as.vector()
    
  },
  FUTURE = T,
  future.seed = NULL,
  .fname = "time") %>% 
  
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> s_pdsi

```

Mapping a random date:
```{r}

s_pdsi[is.na(adrop(vds[,,,1]))] <- NA

sample(date_vector, 1) -> d

ggplot() +
  geom_stars(data = s_pdsi %>% 
               filter(time == d) %>% 
               adrop() %>% 
               mutate(X = case_when(X < -5 ~ -5,
                                    X > 5 ~ 5,
                                    TRUE ~ X))) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80",
                                  name = "Index") +
  coord_fixed() +
  labs(subtitle = d)

```


### 4.1. Temporal correlation
Again, we don't see much improvement. Similar spatial patterns and magnitudes of correlation as when using Thornthwaite and Hargraves. 
```{r}

func_t_cor_map(s_pdsi, vds) -> cor_map

ggplot() +
  geom_stars(data = cor_map) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80", 
                                  limits = c(-0.95,0.95)) +
  coord_fixed()

```

If I randomly choose a pixel with HIGH correlation:
```{r}
func_ts_comparison(0.8)
```

And one with LOW correlation:
```{r}
func_ts_comparison(0.1) 
```

# 1.2. Spatial correlation
Here we see a drastic decline of spatial correlations over time. 
```{r}

c(s_pdsi, vds, along = list(foo = c("era5", "vds"))) %>% 
  split("foo") %>% 
  as_tibble() %>% 
  group_by(time) %>% 
  summarize(cor = cor(era5, vds, use = "complete.obs")) -> t_cor

t_cor %>%   
  ggplot(aes(x = time, y = cor)) +
  geom_point() +
  geom_smooth()

```

If I randomly choose a date with HIGH-ish correlations:
```{r}
func_sp_comparison(0.6)
```

And one with LOW correlations:
```{r}
func_sp_comparison(0.1)
```






