---
title: "scPDSI tests"
author: "Carlos Dobler"
output: github_document
---

Here I show results of a series of test runs for calculating scPDSI with the R package `scpdsi`. All tests were ran for a region centered on ~~Mexico~~ Central Europe to reduce processing times. Of the variables that scPDSI asks for, the only ones that would vary were monthly precipitation and monthly potential evapotranspiration (PET). AWC remained constant (100 mm), and the start and end dates for calibration included the whole period. I used four PET formulations: one uses the "raw" potential evaporation from ERA5 Reanalysis data (no transpiration); the other three were calculated with the R package `SPEI` using various ERA5 Reanalysis variables. 


### Sections:
scPDSI calculated:     
[1. ...with Potential Evaporation (ERA5 data)](#1-with-era5s-potential-evaporation)           
[2. ...with Thornthwaite's formulation](#2-with-thornthwaite)    
[3. ...with Hargreaves' formulation](#3-with-hargreaves)    
[4. ...with Penman-Monteith's formulation](#4-with-penman)    
NEW!!!:    
[5. Penman + elevation](#5-penman-elevation)    
[6. Penman + elevation + AWC](#6-penman-elevation-awc)    


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

source(here::here("scripts/00-set-up.R"))

library(SPEI)
library(scPDSI)
library(colorspace)

plan(multicore, workers = 6)

# load all vars
source(here::here("scripts", "load_variables.R"))

# functions

func_t_cor_map <- function(era_map, vds_map){
  
  c(era_map, vds_map, along = 3) %>% 
  st_apply(c(1,2), function(x){
    
    cor(x[1:504], x[505:1008])
    
  }, 
  FUTURE = T)
}

# ***

func_ts_comparison <- function(thres){
  
  cor_map %>% 
    as_tibble() %>% 
    filter(near(X, thres, 0.01)) %>% 
    slice_sample(n = 1) %>% 
    {c(.$longitude, .$latitude)} -> coords
  
  vds %>%
    as_tibble() %>% 
    filter(longitude == coords[1],
           latitude == coords[2]) %>% 
    pull(scpdsi) -> vect_vds

  s_pdsi %>% 
    as_tibble() %>% 
    filter(longitude == coords[1],
           latitude == coords[2]) %>% 
    pull(X) -> vect_era

  tibble(vds = vect_vds,
         era5 = vect_era,
         time = date_vector) %>% 
    pivot_longer(-time, names_to = "src", values_to = "pdsi") %>% 
    
    ggplot(aes(x = time, y = pdsi, color = src)) +
    geom_line()
  
}

# ***

func_sp_comparison <- function(thres){
  
  t_cor %>% 
    as_tibble() %>% 
    filter(near(cor, thres, 0.01)) %>% 
    slice_sample(n = 1) %>% 
    pull(time) -> t_time
  
  s_pdsi %>%
    filter(time == t_time) %>% 
    adrop() %>% 
    as_tibble() -> m_1
  
  vds %>%
    filter(time == t_time) %>% 
    adrop() %>% 
    as_tibble() -> m_2
  
  left_join(m_1, m_2, by = c("longitude", "latitude")) %>%
    rename(era5 = 3,
           vds = 4) %>% 
    pivot_longer(3:4, names_to = "src", values_to = "pdsi") %>% 
    mutate(pdsi = case_when(pdsi < -5 ~ -5,
                            pdsi > 5 ~ 5,
                            TRUE ~ pdsi)) %>% 
    ggplot(aes(x = longitude, y = latitude, fill = pdsi)) +
    geom_raster() +
    scale_fill_continuous_diverging(palette = "Blue-Red", rev = T, na.value = "grey80") +
    facet_wrap(~src, ncol = 2) +
    coord_equal() +
    labs(subtitle = t_time)
  
}


```


## 1. ...with ERA5's potential evaporation

```{r}

c(var_pr, var_potevap, along = 3) %>% 
  st_apply(c(1,2), function(x){
    
    pdsi(P = x[1:504], 
         PE = x[505:1008], 
         sc = T)$X %>% as.vector()
    
  },
  FUTURE = T,
  future.seed = NULL,
  .fname = "time") %>% 
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> s_pdsi

c(s_pdsi, mask) %>% 
  mutate(X = ifelse(is.na(m), NA, X)) %>% 
  select(X) -> s_pdsi

```

Mapping a random date:
```{r}

sample(date_vector, 1) -> d

ggplot() +
  geom_stars(data = s_pdsi %>% 
               filter(time == d) %>% 
               adrop() %>% 
               mutate(X = case_when(X < -5 ~ -5,
                                    X > 5 ~ 5,
                                    TRUE ~ X))) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80",
                                  name = "PDSI"
                                  ) +
  coord_fixed() +
  labs(subtitle = d)

```

### 1.1. Temporal correlation
The following figure correlates my results against Van der Schrier's PDSI time series on a per-pixel basis:

```{r}

func_t_cor_map(s_pdsi, vds) -> cor_map

ggplot() +
  geom_stars(data = cor_map) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80", 
                                  limits = c(-0.95,0.95),
                                  name = "r") +
  coord_fixed()

```

If I randomly choose a pixel with a *high* correlation coefficient, the time series look like this:

```{r rand_ts_1_1, fig.height=2.5, fig.width=7}
func_ts_comparison(0.8)
```

If I choose one with a LOW correlation coefficient:
```{r rand_ts_1_2, fig.height=2.5, fig.width=7}
func_ts_comparison(0.1)
```


### 1.2. Spatial correlation
The following figure shows correlation coefficients between my results and Van der Schrier's PDSI at a month level (i.e. my resulting map at _t_n vs. VDS resulting map at _t_n):
```{r sp_1, fig.height=2.5, fig.width=7}

c(s_pdsi, vds, along = list(foo = c("vds", "era5"))) %>% 
  split("foo") %>% 
  as_tibble() %>% 
  group_by(time) %>% 
  summarize(cor = cor(era5, vds, use = "complete.obs")) -> t_cor 

t_cor %>%   
  ggplot(aes(x = time, y = cor)) +
  geom_point() +
  geom_smooth()

```

If I randomly choose a date where correlation was *high-ish*, the two maps look like this:
```{r rand_sp_1_1, fig.width=10, fig.height=3.3}
func_sp_comparison(0.65)
```

If I choose one where correlation is *low*:
```{r rand_sp_1_2, fig.width=10, fig.height=3.3}
func_sp_comparison(0.1)
```



## 2. ...with Thornthwaite
Thornthwaite's PET formulation uses two variables: average temperature and latitude.
```{r}

# calculate pet
var_tave %>% # average temp
  split("time") %>% 
  c(var_lat) %>% # lat
  merge() %>% 
  
  st_apply(c(1,2), function(x){
    
    tail(x, 1) -> lat
    x[1:504] -> x_ts
    
    thornthwaite(Tave = x_ts, 
                 lat = lat)
    
  }, 
  FUTURE = T,
  .fname = "time") %>% 
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> pet

# calculate pdsi
c(var_pr, pet, along = 3) %>%
  st_apply(c(1,2), function(x){

    pdsi(P = x[1:504],
         PE = x[505:1008],
         sc = T)$X %>% as.vector()

  },
  FUTURE = T,
  future.seed = NULL,
  .fname = "time") %>%

  st_set_dimensions("time", values = date_vector) %>%
  aperm(c(2,3,1)) -> s_pdsi

c(s_pdsi, mask) %>% 
  mutate(X = ifelse(is.na(m), NA, X)) %>% 
  select(X) -> s_pdsi

```

Mapping a random date:
```{r}

sample(date_vector, 1) -> d

ggplot() +
  geom_stars(data = s_pdsi %>% 
               filter(time == d) %>% 
               adrop() %>% 
               mutate(X = case_when(X < -5 ~ -5,
                                    X > 5 ~ 5,
                                    TRUE ~ X))) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80",
                                  name = "PDSI") +
  coord_fixed() +
  labs(subtitle = d)

```


### 2.1. Temporal correlation
When correlating my results using Thornthwaite vs Van der Schrier's, we see a drastic improvement over previous results. All cells show positive correlations, and with high a coefficient:
```{r}

func_t_cor_map(s_pdsi, vds) -> cor_map

ggplot() +
  geom_stars(data = cor_map) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80", 
                                  limits = c(-0.95,0.95),
                                  name = "r") +
  coord_fixed()

```

If I randomly choose a pixel with *high* correlation, the time series look like this:
```{r rand_ts_2_1, fig.height=2.5, fig.width=7}
func_ts_comparison(0.8) 
```

And with *low* correlation:
```{r rand_ts_2_2, fig.height=2.5, fig.width=7}
func_ts_comparison(0.15)
```

### 2.2. Spatial correlation
Spatially, correlation between my results and Van der Schrier's fluctuates around ~0.1 and ~0.7:
```{r sp_2, fig.height=2.5, fig.width=7}

c(s_pdsi, vds, along = list(foo = c("vds", "era5"))) %>% 
  split("foo") %>% 
  as_tibble() %>% 
  group_by(time) %>% 
  summarize(cor = cor(era5, vds, use = "complete.obs")) -> t_cor 

t_cor %>%   
  ggplot(aes(x = time, y = cor)) +
  geom_point() +
  geom_smooth()

```

If I randomly choose a date when correlation was *high*:
```{r rand_sp_2_1, fig.width=10, fig.height=3.3}
func_sp_comparison(0.65)
```
And a date when correlation was *low*:
```{r rand_sp_2_2, fig.width=10, fig.height=3.3}
func_sp_comparison(0.1)
```



## 3. ...with Hargreaves
Hargraves' PET formulation uses three variables: maximum temperature, minimum temperature, and radiation.

```{r}

# pet
c(var_ra, var_tmax, var_tmin, along = 3) %>%
  st_apply(c(1,2), function(x){
    
    hargreaves(Tmin = x[1009:1512],
               Tmax = x[505:1008],
               Ra = x[1:504],
               na.rm = T)
    
  },
  FUTURE = T,
  .fname = "time") %>% 
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> pet

# pdsi
c(var_pr, pet, along = 3) %>% 
  st_apply(c(1,2), function(x){
    
    pdsi(P = x[1:504], 
         PE = x[505:1008], 
         sc = T)$X %>% as.vector()
    
  },
  FUTURE = T,
  future.seed = NULL,
  .fname = "time") %>% 
  
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> s_pdsi

c(s_pdsi, mask) %>% 
  mutate(X = ifelse(is.na(m), NA, X)) %>% 
  select(X) -> s_pdsi

```

Mapping a random date:
```{r}

sample(date_vector, 1) -> d

ggplot() +
  geom_stars(data = s_pdsi %>% 
               filter(time == d) %>% 
               adrop() %>% 
               mutate(X = case_when(X < -5 ~ -5,
                                    X > 5 ~ 5,
                                    TRUE ~ X))) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80",
                                  name = "PDSI") +
  coord_fixed() +
  labs(subtitle = d)

```


### 3.1. Temporal correlation
Hargraves seem to do worse than Thornthwaite on a per-pixel basis.
```{r}

func_t_cor_map(s_pdsi, vds) -> cor_map

ggplot() +
  geom_stars(data = cor_map) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80", 
                                  limits = c(-0.95,0.95),
                                  name = "r") +
  coord_fixed()

```

If I randomly choose a pixel with *high* correlation:
```{r rand_ts_3_1, fig.height=2.5, fig.width=7}
func_ts_comparison(0.8)
```

And one with *low* correlation:
```{r rand_ts_3_2, fig.height=2.5, fig.width=7}
func_ts_comparison(0.1) 
```

### 3.2. Spatial correlation
Interestingly, the spatial correlation between my results and Van der Schrier's show an "arching" trend over time: 
```{r sp_3, fig.height=2.5, fig.width=7}

c(s_pdsi, vds, along = list(foo = c("vds", "era5"))) %>% 
  split("foo") %>% 
  as_tibble() %>% 
  group_by(time) %>% 
  summarize(cor = cor(era5, vds, use = "complete.obs")) -> t_cor 

t_cor %>%   
  ggplot(aes(x = time, y = cor)) +
  geom_point() +
  geom_smooth()

```

If I randomly choose a date with a *high-ish* correlation it looks like this:
```{r rand_sp_3_1, fig.width=10, fig.height=3.3}
func_sp_comparison(0.65)
```

And one with *low* correlation:
```{r rand_sp_3_2, fig.width=10, fig.height=3.3}
func_sp_comparison(0.1)
```



## 4. ...with Penman
Penman-Monteith's PET formulation can include several variables; some are used to derive others. In this case, I used: maximum temperature, minimum temperature, wind speed (it should be at 2m but I used at 10m), external radiance (I used "top of atmosphere"), incoming radiance, dewpoint temperature (to derive vapor pressure), and surface pressure.  

```{r}

# pet
source(here::here("scripts", "penman_mod.R"))

c(var_tmin, var_tmax, var_wind, var_ra, var_rs, var_dewpoint, var_pressure, along = 3) -> stack

stack %>% 
  st_apply(c(1,2), function(x){
    
    penman_mod(Tmin = x[1:504],
               Tmax = x[505:1008],
               U2 = x[1009:1512],
               Ra = x[1513:2016],
               Rs = x[2017:2520],
               Tdew = x[2521:3024],
               P = x[3025:3528])
    
  },
  FUTURE = T,
  .fname = "time") %>% 
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> pet

# scpdsi
c(var_pr, pet, along = 3) %>% 
  st_apply(c(1,2), function(x){
    
    pdsi(P = x[1:504], 
         PE = x[505:1008], 
         sc = T)$X %>% as.vector()
    
  },
  FUTURE = T,
  future.seed = NULL,
  .fname = "time") %>% 
  
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> s_pdsi

c(s_pdsi, mask) %>% 
  mutate(X = ifelse(is.na(m), NA, X)) %>% 
  select(X) -> s_pdsi

```

Mapping a random date:
```{r}

sample(date_vector, 1) -> d

ggplot() +
  geom_stars(data = s_pdsi %>% 
               filter(time == d) %>% 
               adrop() %>% 
               mutate(X = case_when(X < -5 ~ -5,
                                    X > 5 ~ 5,
                                    TRUE ~ X))) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80",
                                  name = "Index") +
  coord_fixed() +
  labs(subtitle = d)

```


### 4.1. Temporal correlation
Again, we don't see much improvement over Thornthwaite's. Similar spatial patterns and magnitudes of correlation as when using Hargraves. 
```{r}

func_t_cor_map(s_pdsi, vds) -> cor_map

ggplot() +
  geom_stars(data = cor_map) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80", 
                                  limits = c(-0.95,0.95),
                                  name = "r") +
  coord_fixed()

```

If I randomly choose a pixel with *high* correlation:
```{r rand_ts_4_1, fig.height=2.5, fig.width=7}
func_ts_comparison(0.8)
```

And one with *low* correlation:
```{r rand_ts_4_2, fig.height=2.5, fig.width=7}
func_ts_comparison(0.1) 
```

### 4.2. Spatial correlation
Similarly, on a spatial basis, correlations display and "arched" trend over time.
```{r sp_4, fig.height=2.5, fig.width=7}

c(s_pdsi, vds, along = list(foo = c("era5", "vds"))) %>% 
  split("foo") %>% 
  as_tibble() %>% 
  group_by(time) %>% 
  summarize(cor = cor(era5, vds, use = "complete.obs")) -> t_cor

t_cor %>%   
  ggplot(aes(x = time, y = cor)) +
  geom_point() +
  geom_smooth()

```

If I randomly choose a date with *high-ish* correlations:
```{r rand_sp_4_1, fig.width=10, fig.height=3.3}
func_sp_comparison(0.65)
```

And one with *low* correlations:
```{r rand_sp_4_2, fig.width=10, fig.height=3.3}
func_sp_comparison(0.1)
```

## 5. Penman + elevation

Elevation is an additional variable that can be used to estimate PET under Penman's formulation. I used ERA5 geopotential height divided by the Earth's gravitational acceleration, g (= 9.80665 m s-2).

```{r}

# pet
abind(stack[[1]], var_z[[1]], along = 3) -> stack_array
dim(stack_array) <- c(x = 78, y = 60, time = 3529)

st_as_stars(stack_array) %>% 
  st_apply(c(1,2), function(x){
    
    penman_mod(Tmin = x[1:504],
               Tmax = x[505:1008],
               U2 = x[1009:1512],
               Ra = x[1513:2016],
               Rs = x[2017:2520],
               Tdew = x[2521:3024],
               P = x[3025:3528],
               z = x[3529])
    
  },
  FUTURE = T,
  .fname = "time") %>% 
  aperm(c(2,3,1)) -> pet

st_dimensions(pet) <- st_dimensions(var_tmin)

# scpdsi
c(var_pr, pet, along = 3) %>% 
  st_apply(c(1,2), function(x){
    
    pdsi(P = x[1:504], 
         PE = x[505:1008], 
         sc = T)$X %>% as.vector()
    
  },
  FUTURE = T,
  future.seed = NULL,
  .fname = "time") %>% 
  
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> s_pdsi

c(s_pdsi, mask) %>% 
  mutate(X = ifelse(is.na(m), NA, X)) %>% 
  select(X) -> s_pdsi

```

### 5.1. Temporal correlation
Elevation doesn't seem to improve things on a per-pixel basis... 
```{r}

func_t_cor_map(s_pdsi, vds) -> cor_map

ggplot() +
  geom_stars(data = cor_map) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80", 
                                  limits = c(-0.95,0.95),
                                  name = "r") +
  coord_fixed()

```

### 5.2. Spatial correlation
And spatially, variability in correlation look very similar...
```{r sp_5, fig.height=2.5, fig.width=7}

c(s_pdsi, vds, along = list(foo = c("era5", "vds"))) %>% 
  split("foo") %>% 
  as_tibble() %>% 
  group_by(time) %>% 
  summarize(cor = cor(era5, vds, use = "complete.obs")) -> t_cor

t_cor %>%   
  ggplot(aes(x = time, y = cor)) +
  geom_point() +
  geom_smooth()

```

## 6. Penman + elevation + AWC

AWC is a variable required in the calculation of PDSI. I found several AWC datasets, but all of them are quite complicated to obtain and process (either antique, bizarre, or proprietary data formats). Before going through the pain of dealing with this, I ran an experiment to see to what extent AWC could improve results. The experiment consists of altering iteratively AWC values for each pixel and choose the one that gets me "the best" PDSI (i.e. the one with the highest correlation coefficient with Van der Schrier's data). While in the former runs AWC was kept constant (100 mm), here it can have values from 5 to 215.

```{r}

func_x <- function(x){
  
  if(sum(!is.na(x[1009:1512])) == 0){
      rep(NA, 506)
     } else{
       
       map_dfr(seq(5, 230, 30), function(awc_i){
         
         pdsi(P = x[1:504], 
              PE = x[505:1008], 
              AWC = awc_i,
              sc = T)$X %>% as.vector() -> pdsi_result
         
         tibble(pdsi = pdsi_result,
                awc = awc_i)
         
       }) -> tb_awc
       
       tb_awc %>% 
         group_by(awc) %>% 
         summarize(r = cor(pdsi, x[1009:1512])) %>% 
         filter(r == max(r, na.rm = T)) %>% 
         slice_sample(n = 1) -> tb_cor
       
       tb_awc %>% 
         filter(awc == tb_cor$awc) %>% 
         pull(pdsi) -> vect_pdsi
       
       # vect_pdsi
       c(vect_pdsi, tb_cor$awc, tb_cor$r)
  
     }
}

# ******

c(var_pr, pet, vds, along = 3) %>%
  
  st_apply(c(1,2),
           func_x,
           FUTURE = T,
           future.seed = NULL,
           .fname = "time") -> s_pdsi

s_pdsi %>% 
  slice(time, 505) -> s_awc

s_pdsi %>% 
  slice(time, 506) -> s_cor

s_pdsi %>%
  slice(time, 1:504) %>% 
  st_set_dimensions("time", values = date_vector) %>% 
  aperm(c(2,3,1)) -> s_pdsi


```

This is how the "artificial" AWC layer looks like. We can see that much of central Europe was run with low AWC values (< 50):
```{r}

ggplot() +
  geom_stars(data = s_awc) +
  scale_fill_continuous_sequential(palette = "Viridis", 
                                   rev = T, 
                                   na.value = "grey80",
                                   name = "mm") +
  coord_fixed()

```


### 6.1. Temporal correlation
Unfortunately, the addition of AWC does not seem to improve results:
```{r}

func_t_cor_map(s_pdsi, vds) -> cor_map

ggplot() +
  geom_stars(data = cor_map) +
  scale_fill_continuous_diverging(palette = "Blue-Red", 
                                  rev = T, 
                                  na.value = "grey80", 
                                  limits = c(-0.95,0.95),
                                  name = "r") +
  coord_fixed()

```

### 6.2. Spatial correlation
And same spatially:
```{r sp_6, fig.height=2.5, fig.width=7}

c(s_pdsi, vds, along = list(foo = c("era5", "vds"))) %>% 
  split("foo") %>% 
  as_tibble() %>% 
  group_by(time) %>% 
  summarize(cor = cor(era5, vds, use = "complete.obs")) -> t_cor

t_cor %>%   
  ggplot(aes(x = time, y = cor)) +
  geom_point() +
  geom_smooth()

```



