---
title: "REMO vs ERA in a 1 degree world: PET & Precip"
author: "Carlos Dobler"
date: "2/25/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(colorspace)

source(here::here("scripts/00-set-up.R"))

c(st_point(c(-11, 65)),
  st_point(c(39, 32))) %>% 
  st_bbox() %>% 
  st_set_crs(4326) -> lim

source(here::here("scripts/remo.R"))

```

Mean monthly PET:

```{r}

r_models_pet_ann %>% 
  st_warp(e_var_pet_ann) %>% 
  split("model") %>% 
  {c(e_var_pet_ann, .)} -> all_pet_ann

all_pet_ann[is.na(r_land_rast)] <- NA
  
all_pet_ann %>% 
  merge(name = "model") %>% 
  {
    ggplot() +
      geom_stars(data = .) +
      facet_wrap(~model, ncol = 2) +
      scale_fill_continuous_sequential("Plasma", name = "mm", rev = F) +
      coord_fixed()
      
  }

```

Differences in mean monthly PET between REMO and ERA:

```{r}

all_pet_ann %>% 
  mutate(ERA_Had = ERA - Had,
         ERA_MPI = ERA - MPI,
         ERA_Nor = ERA - Nor,
         ERA_ENS = ERA - ENS) %>% 
  select(-c(1:5)) %>% 
  merge(name = "model") %>% 
  st_set_dimensions("model", values = str_replace(st_get_dimension_values(., "model"), "_", " - ")) %>% 
  
  {
    ggplot() +
      geom_stars(data = .) +
      facet_wrap(~model, ncol = 2) +
      scale_fill_continuous_divergingx("Spectral", name = "mm", rev = F) +
      coord_fixed()
      
  }

```

Mean monthly precipitation:

```{r}

r_models_pr_ann %>% 
  st_warp(e_var_pr_ann) %>% 
  split("model") %>% 
  {c(e_var_pr_ann, .)} -> all_pr_ann

all_pr_ann[is.na(r_land_rast)] <- NA
  
all_pr_ann %>% 
  merge(name = "model") %>% 
  {
    ggplot() +
      geom_stars(data = .) +
      facet_wrap(~model, ncol = 2) +
      scale_fill_continuous_sequential("Plasma", name = "mm", rev = F) +
      coord_fixed()
      
  }

```

Differences in mean monthly precipitation between REMO and ERA:

```{r}

all_pr_ann %>% 
  mutate(ERA_Had = ERA - Had,
         ERA_MPI = ERA - MPI,
         ERA_Nor = ERA - Nor,
         ERA_ENS = ERA - ENS) %>% 
  select(-c(1:5)) %>% 
  merge(name = "model") %>% 
  st_set_dimensions("model", values = str_replace(st_get_dimension_values(., "model"), "_", " - ")) %>% 
  
  {
    ggplot() +
      geom_stars(data = .) +
      facet_wrap(~model, ncol = 2) +
      scale_fill_continuous_divergingx("Spectral", name = "mm", rev = F) +
      coord_fixed()
      
  }


```

Mean count of months/year with SPEI-3 < -1.5:

```{r}

r_models_spei_ann %>% 
  st_warp(e_var_spei_ann) %>% 
  split("model") %>% 
  {c(e_var_spei_ann, .)} -> all_spei_ann

all_spei_ann[is.na(r_land_rast)] <- NA
  
all_spei_ann %>% 
  merge(name = "model") %>% 
  {
    ggplot() +
      geom_stars(data = .) +
      facet_wrap(~model, ncol = 2) +
      scale_fill_continuous_sequential("Plasma", name = "count of months", rev = F) +
      coord_fixed()
      
  }

```

Differences in count of months/year with SPEI-3 < -1.5 between REMO and ERA:

```{r}

all_spei_ann %>% 
  mutate(ERA_Had = ERA - Had,
         ERA_MPI = ERA - MPI,
         ERA_Nor = ERA - Nor,
         ERA_ENS = ERA - ENS) %>% 
  select(-c(1:5)) %>% 
  merge(name = "model") %>% 
  st_set_dimensions("model", values = str_replace(st_get_dimension_values(., "model"), "_", " - ")) %>% 
  
  {
    ggplot() +
      geom_stars(data = .) +
      facet_wrap(~model, ncol = 2) +
      scale_fill_continuous_divergingx("Spectral", name = "diff.\n(count of months)", rev = F) +
      coord_fixed()
      
  }

```





Mean seasonal cycle for PET:

```{r}

r_models_pet_msc %>%
  map(st_warp, e_var_pet_msc) %>% 
  do.call(c, .) %>% 
  {c(e_var_pet_msc, .)} %>% 
  setNames(c("ERA", "Had", "MPI", "Nor", "ENS")) -> all_pet_msc

all_pet_msc[is.na(r_land_rast)] <- NA
  
all_pet_msc %>% 
  merge(name = "model") %>%
  slice(month, c(1,4,7,10)) %>% 
  {
    ggplot() +
      geom_stars(data = .) +
      facet_grid(model~month) +
      scale_fill_continuous_sequential("Plasma", name = "mm", rev = F) +
      coord_fixed()
  }

```

Differences in the mean seasonal cycle between REMO and ERA:

```{r}

all_pet_msc %>% 
  mutate(ERA_Had = ERA - Had,
         ERA_MPI = ERA - MPI,
         ERA_Nor = ERA - Nor,
         ERA_ENS = ERA - ENS) %>% 
  select(-c(1:5)) %>% 
  merge(name = "model") %>% 
  st_set_dimensions("model", values = str_replace(st_get_dimension_values(., "model"), "_", " - ")) %>% 
  slice(month, c(1,4,7,10)) %>% 
  
  {
    ggplot() +
      geom_stars(data = .) +
      facet_grid(model~month) +
      scale_fill_continuous_divergingx("Spectral", name = "mm", rev = F) +
      coord_fixed()
      
  }

```

Mean seasonal cycle for precipitation:

```{r}
r_models_pr_msc %>%
  map(st_warp, e_var_pr_msc) %>% 
  do.call(c, .) %>% 
  {c(e_var_pr_msc, .)} %>% 
  setNames(c("ERA", "Had", "MPI", "Nor", "ENS")) -> all_pr_msc

all_pr_msc[is.na(r_land_rast)] <- NA
  
all_pr_msc %>% 
  merge(name = "model") %>%
  slice(month, c(1,4,7,10)) %>% 
  {
    ggplot() +
      geom_stars(data = .) +
      facet_grid(model~month) +
      scale_fill_continuous_sequential("Plasma", name = "mm", rev = F) +
      coord_fixed()
      
  }
```

Differences in precipitation's mean seasonal cycle between REMO and ERA 

```{r}

all_pr_msc %>% 
  mutate(ERA_Had = ERA - Had,
         ERA_MPI = ERA - MPI,
         ERA_Nor = ERA - Nor,
         ERA_ENS = ERA - ENS) %>% 
  select(-c(1:5)) %>% 
  merge(name = "model") %>% 
  st_set_dimensions("model", values = str_replace(st_get_dimension_values(., "model"), "_", " - ")) %>% 
  slice(month, c(1,4,7,10)) %>% 
  
  {
    ggplot() +
      geom_stars(data = .) +
      facet_grid(model~month) +
      scale_fill_continuous_divergingx("Spectral", name = "mm", rev = F) +
      coord_fixed()
      
  }

```

Mean seasonal cycle for count of months/year with SPEI-3 < -1.5:

```{r}

r_models_spei_msc %>%
  map(st_warp, e_var_spei_msc) %>% 
  do.call(c, .) %>% 
  {c(e_var_spei_msc, .)} %>% 
  setNames(c("ERA", "Had", "MPI", "Nor", "ENS")) -> all_spei_msc

all_spei_msc[is.na(r_land_rast)] <- NA
  
all_spei_msc %>% 
  merge(name = "model") %>%
  slice(month, c(1,4,7,10)) %>% 
  {
    ggplot() +
      geom_stars(data = .) +
      facet_grid(model~month) +
      scale_fill_binned_sequential("Plasma", name = "count of months", rev = F) +
      coord_fixed()
  }

```

Differences in the mean seasonal cycle for count of months/year with SPEI-3 < -1.5 between REMO and ERA:

```{r}

all_spei_msc %>% 
  mutate(ERA_Had = ERA - Had,
         ERA_MPI = ERA - MPI,
         ERA_Nor = ERA - Nor,
         ERA_ENS = ERA - ENS) %>% 
  select(-c(1:5)) %>% 
  merge(name = "model") %>% 
  st_set_dimensions("model", values = str_replace(st_get_dimension_values(., "model"), "_", " - ")) %>% 
  slice(month, c(1,4,7,10)) %>% 
  
  {
    ggplot() +
      geom_stars(data = .) +
      facet_grid(model~month) +
      scale_fill_continuous_divergingx("Spectral", name = "count of months", rev = F) +
      coord_fixed()
      
  }

```


