---
title: "Report 3: REMO vs ERA vs vdS - 1 degree world"
author: "Carlos Dobler"
output:
  github_document:
    toc: true
    toc_depth: 2
---

This 3rd report compares scPDSI obtained from REMO2015, ERA5, and van der Schrier et al. **in a +1 degree world**. Same as the previous report, scPDSI from REMO2015 was calculated from scratch, and likewise, the calibration period includes the whole temporal period (1979-2020).

Comparison methods used here differ from those of previous reports since data sources sliced to correspond to 1 degree do not share the same temporality. Thus, I compared spatial patterns of 20th and 80th percentiles between data sources both annually and seasonally.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      out.width = '80%')

source(here::here("scripts/00-set-up.R"))

library(colorspace)

plan(multicore, workers = 6)

# load all vars

readRDS(here::here("output/r_models_pdsi_1deg.rds")) -> r_models_pdsi
readRDS(here::here("output/e_pdsi_pm_1979_2020.rds")) -> e_pdsi
e_pdsi %>% 
  filter(year(time) >= 2000,
         year(time) <= 2020) -> e_pdsi
readRDS(here::here("output/v_pdsi_1979_2020.rds")) -> v_pdsi
v_pdsi %>% 
  filter(year(time) >= 2000,
         year(time) <= 2020) -> v_pdsi
  

# land mask
# warped at ERA resolution

st_read("~/bucket_mine/misc_data/ne_50m_land/ne_50m_land.shp") %>%
  mutate(a = 1) %>%
  select(a) -> land

land %>% 
  st_rasterize(st_as_stars(st_bbox(), dx = 0.25, dy = 0.25, values = NA)) -> land_rast

st_warp(land_rast %>%
          st_set_dimensions(which = c(1,2),
                            names = c("longitude", "latitude")),
        e_pdsi %>% slice(time,1)) -> land_rast

# mask

r_models_pdsi %>%
  map(st_warp, land_rast) %>% 
  map(function(s){
    s[is.na(land_rast)] <- NA
    return(s)
  }) -> r_models_pdsi

v_pdsi[is.na(land_rast)] <- NA
e_pdsi[is.na(land_rast)] <- NA

# # merge

models <- c("Had", "MPI", "Nor", "ENS", "ERA", "vdS")

```

## 1. Annual patterns

### 1.1. 20th percentile
The following maps show the 20th percentile of scPDSI of the different data sources. Compared to ERA5 and vdS, REMO models tend to depict less severe levels of extreme drought.

```{r fig_1_1_1, fig.height=4.5}

append(r_models_pdsi, list(ERA = e_pdsi, vdS = v_pdsi)) %>% 
  imap(function(s, i){
    
    s %>% 
      st_apply(c(1,2),
               function(x){
                 
                 quantile(x, probs = c(0.2, 0.8), na.rm = T)
                 
               }, 
               FUTURE = T,
               .fname = "q") %>% 
      
      setNames(i)
    
  }) %>% 
  unname() %>% 
  do.call(c, .) -> all_pdsi_percs

all_pdsi_percs %>%
  select(1:3) %>% 
  merge(name = "model") %>% 
  st_apply(c(1,2,3), mean, na.rm = T, .fname = "ENS", FUTURE = T) %>% 
  {c(all_pdsi_percs, .)} %>% 
  select(c(1:3,6,4,5)) %>% 
  merge(name = "model") %>% 
  mutate(X = case_when(X < -5 ~ -5,
                       X > 5 ~ 5,
                       TRUE ~ X)) -> all_pdsi_percs

all_pdsi_percs %>% 
  slice(q, 1) %>%
  
  {
    ggplot() +
      geom_stars(data = .) +
      facet_wrap(~model) +
      scale_fill_binned_divergingx(palette = "RdYlBu", 
                                   rev = F, 
                                   na.value = "transparent",
                                   name = "PDSI",
                                   limits = c(-5,5),
                                   n.breaks = 9) +
      geom_polygon(data = land %>% st_crop(land_rast) %>% as("Spatial") %>% fortify(), 
                 fill = NA,
                 color = "black",
                 aes(x = long, y = lat, group = group)) +
      
      guides(fill = guide_colorsteps(barwidth = 15, barheight = 0.5)) +
      theme(legend.position = "bottom") +
      
      labs(x = NULL,
           y = NULL) +
      
      coord_equal()

  }

```

Differences can better be appreciated in the following figure, which shows pair-wise comparisons (subtractions) between data sources. In general, REMO models tend to show lower levels of extreme drought than ERA5. This is also true when compared to vdS, but only for central and western Europe. In eastern Europe/Russia, REMO models tend to be show higher levels of drought than vdS. ERA and vdS show considerable discrepancies in the Scandinavian peninsula and eastern Europe/Russia.

```{r fig_1_1_2, fig.height=5}

tibble(ds1 = seq_along(models)) %>%
  expand(ds1, ds2 = ds1) %>% 
  filter(ds1 < ds2) -> ds

ds %>% 
  filter(ds2 %in% c(5,6)) -> ds

pmap(ds, function(ds1, ds2){

  str_glue("{models[ds1]}_vs_{models[ds2]}") -> var_name
  
  all_pdsi_percs %>%
    slice(q, 1) %>%
    split("model") %>% 
    rename("var1" = ds1,
           "var2" = ds2) %>% 
    mutate(dif = var1 - var2) %>% 
    select(dif) %>% 
    setNames(var_name)
    
}) -> tb_comp

tb_comp %>% 
  do.call(c, .) %>% 
  merge(name = "comp") -> mann_comparisons

mann_comparisons %>% 
  st_get_dimension_values("comp") %>% 
  str_replace("_vs_", " - ") %>% 
  {st_set_dimensions(mann_comparisons, "comp", values = .)} -> mann_comparisons

mann_comparisons %>%
  mutate(X = case_when(X > 5 ~ 5,
                       X < -5 ~ -5,
                       TRUE ~ X)) -> mann_comparisons

comp_levels <-  c("Had - ERA",
                  "MPI - ERA",
                  "Nor - ERA",
                  "ENS - ERA",
                  
                  "Had - vdS",
                  "MPI - vdS",
                  "Nor - vdS",
                  "ENS - vdS",
                  
                  "ERA - vdS")

ggplot() +
  geom_stars(data = mann_comparisons) +
  facet_wrap(~comp %>% 
               factor(levels = comp_levels),
             ncol = 4) +
  scale_fill_binned_diverging(palette = "Vik",
                                  rev = T,
                                  na.value = "transparent",
                                  limits = c(-5,5),
                                  n.breaks = 11,
                              name = "difference\n(PDSI)") +
  
  theme(strip.text.x = element_text(size = 7.5)) +
  geom_polygon(data = land %>% st_crop(land_rast) %>% as("Spatial") %>% fortify(), 
                 fill = NA,
                 color = "black",
                 aes(x = long, y = lat, group = group)) +
      
      guides(fill = guide_colorsteps(barwidth = 15, barheight = 0.5)) +
      theme(legend.position = "bottom") +
      
      labs(x = NULL,
           y = NULL) +
      
      coord_equal()

```

### 1.2. 80th percentile
The following maps show the 80th percentile of the different data sources' scPDSI. Compared to ERA5 and vdS, REMO models tend to display wetter conditions. In fact, ERA5 and vdS show considerable areas with negative PDSI, which does not happen in REMO models.

```{r fig_1_2_1, fig.height=4.5}

all_pdsi_percs %>% 
  slice(q, 2) %>%
  
  {
    ggplot() +
      geom_stars(data = .) +
      facet_wrap(~model) +
      scale_fill_binned_divergingx(palette = "RdYlBu", 
                                   rev = F, 
                                   na.value = "transparent",
                                   name = "PDSI",
                                   limits = c(-5,5),
                                   n.breaks = 9) +
      geom_polygon(data = land %>% st_crop(land_rast) %>% as("Spatial") %>% fortify(), 
                 fill = NA,
                 color = "black",
                 aes(x = long, y = lat, group = group)) +
      
      guides(fill = guide_colorsteps(barwidth = 15, barheight = 0.5)) +
      theme(legend.position = "bottom") +
      
      labs(x = NULL,
           y = NULL) +
      
      coord_equal()

  }

```

The following figure shows pair-wise differences between data sources for the 80th percentile. The figure confirms that REMO models present wetter conditions than ERA5 and vdS.

```{r fig_1_2_2, fig.height=5}

pmap(ds, function(ds1, ds2){

  str_glue("{models[ds1]}_vs_{models[ds2]}") -> var_name
  
  all_pdsi_percs %>%
    slice(q, 2) %>%
    split("model") %>% 
    rename("var1" = ds1,
           "var2" = ds2) %>% 
    mutate(dif = var1 - var2) %>% 
    select(dif) %>% 
    setNames(var_name)
    
}) -> tb_comp

tb_comp %>% 
  do.call(c, .) %>% 
  merge(name = "comp") -> mann_comparisons

mann_comparisons %>% 
  st_get_dimension_values("comp") %>% 
  str_replace("_vs_", " - ") %>% 
  {st_set_dimensions(mann_comparisons, "comp", values = .)} -> mann_comparisons

mann_comparisons %>%
  mutate(X = case_when(X > 5 ~ 5,
                       X < -5 ~ -5,
                       TRUE ~ X)) -> mann_comparisons

ggplot() +
  geom_stars(data = mann_comparisons) +
  facet_wrap(~comp %>% 
               factor(levels = comp_levels),
             ncol = 4) +
  scale_fill_binned_diverging(palette = "Vik",
                                  rev = T,
                                  na.value = "transparent",
                                  limits = c(-5,5),
                                  n.breaks = 11,
                              name = "difference\n(PDSI)") +
  
  theme(strip.text.x = element_text(size = 7.5)) +
  geom_polygon(data = land %>% st_crop(land_rast) %>% as("Spatial") %>% fortify(), 
                 fill = NA,
                 color = "black",
                 aes(x = long, y = lat, group = group)) +
      
      guides(fill = guide_colorsteps(barwidth = 15, barheight = 0.5)) +
      theme(legend.position = "bottom") +
      
      labs(x = NULL,
           y = NULL) +
      
      coord_equal()

```

## 2. Seasonal patterns

### 2.1. 20th percentile

The following figure shows maps of the 20th percentile scPDSI for the different data sources over different seasons. Again, it shows that ERA5 and vdS tend to depict drier conditions than REMO models.

```{r fig_2_1_1, fig.height=7.5}

seas <- list(DJF = c(12,1,2), 
             MAM = c(3,4,5),
             JJA = c(6,7,8),
             SON = c(9,10,11))

append(r_models_pdsi, list(ERA = e_pdsi, vdS = v_pdsi)) %>% 
  imap(function(s, i){
    
    imap(seas, function(seass, ii){
      
      s %>% 
        filter(month(time) %in% seass) %>% 
        
        st_apply(c(1,2),
                 function(x){
                   
                   quantile(x, probs = c(0.2, 0.8), na.rm = T)
                   
                 },
                 FUTURE = T,
                 .fname = "q") %>% 
        
        setNames(ii)
      
    }) %>% 
      unname() %>% 
      do.call(c, .) %>% 
      merge(name = "season") %>% 
      setNames(i)
    
  }) %>% 
  unname() %>% 
  do.call(c, .) -> all_pdsi_percs_seas

all_pdsi_percs_seas %>%
  select(1:3) %>%
  merge(name = "model") %>% 
  st_apply(c(1,2,3,4), mean, na.rm = T, .fname = "ENS", FUTURE = T) %>% 
  {c(all_pdsi_percs_seas, .)} %>% 
  select(c(1:3,6,4,5)) %>% 
  merge(name = "model") %>% 
  mutate(X = case_when(X < -5 ~ -5,
                       X > 5 ~ 5,
                       TRUE ~ X)) -> all_pdsi_percs_seas

all_pdsi_percs_seas %>% 
  slice(q, 1) %>%
  
  {
    ggplot() +
      geom_stars(data = .) +
      facet_grid(model~season) +
      scale_fill_binned_divergingx(palette = "RdYlBu", 
                                   rev = F, 
                                   na.value = "transparent",
                                   name = "PDSI",
                                   limits = c(-5,5),
                                   n.breaks = 9) +
      geom_polygon(data = land %>% st_crop(land_rast) %>% as("Spatial") %>% fortify(), 
                 fill = NA,
                 color = "black",
                 aes(x = long, y = lat, group = group)) +
      
      guides(fill = guide_colorsteps(barwidth = 15, barheight = 0.5)) +
      theme(legend.position = "bottom") +
      
      labs(x = NULL,
           y = NULL) +
      
      coord_equal()

  }

```

The following figure shows pair-wise comparisons between data sources (only REMO ensemble). 

```{r fig_2_1_2, fig.height=4.5}

models_s <- models[4:6]

tibble(ds1 = seq_along(models_s)) %>%
  expand(ds1, ds2 = ds1) %>% 
  filter(ds1 < ds2) -> ds

pmap(ds, function(ds1, ds2){

  str_glue("{models_s[ds1]}_vs_{models_s[ds2]}") -> var_name
  
  all_pdsi_percs_seas %>%
    slice(q, 1) %>%
    split("model") %>% 
    select(models_s) %>% 
    rename("var1" = ds1,
           "var2" = ds2) %>% 
    mutate(dif = var1 - var2) %>% 
    select(dif) %>% 
    setNames(var_name)
    
}) -> tb_comp

tb_comp %>% 
  do.call(c, .) %>% 
  merge(name = "comp") -> mann_comparisons

mann_comparisons %>% 
  st_get_dimension_values("comp") %>% 
  str_replace("_vs_", " - ") %>% 
  {st_set_dimensions(mann_comparisons, "comp", values = .)} -> mann_comparisons

mann_comparisons %>%
  mutate(X = case_when(X > 5 ~ 5,
                       X < -5 ~ -5,
                       TRUE ~ X)) -> mann_comparisons

comp_levels <-  c("ENS - ERA",
                  "ENS - vdS",
                  "ERA - vdS")

ggplot() +
  geom_stars(data = mann_comparisons) +
  facet_grid(comp %>% factor(levels = comp_levels) ~ season) +
  scale_fill_binned_diverging(palette = "Vik",
                              rev = T,
                              na.value = "transparent",
                              limits = c(-5,5),
                              n.breaks = 11,
                              name = "difference\n(PDSI)") +
  
  theme(strip.text.x = element_text(size = 7.5)) +
  geom_polygon(data = land %>% st_crop(land_rast) %>% as("Spatial") %>% fortify(), 
               fill = NA,
               color = "black",
               aes(x = long, y = lat, group = group)) +
  
  guides(fill = guide_colorsteps(barwidth = 15, barheight = 0.5)) +
  theme(legend.position = "bottom") +
  
  labs(x = NULL,
       y = NULL) +
  
  coord_equal()

```

### 2.2. 80th percentile

The following figure shows maps of the 80th percentile scPDSI for the different data sources over different seasons. As in the annual patterns, it shows that ERA5 and vdS tend to depict drier conditions than REMO models, again, with considerably sized regions displaying negative PDSI values in the former.

```{r fig_2_2_1, fig.height=7.5}

all_pdsi_percs_seas %>% 
  slice(q, 2) %>%
  
  {
    ggplot() +
      geom_stars(data = .) +
      facet_grid(model~season) +
      scale_fill_binned_divergingx(palette = "RdYlBu", 
                                   rev = F, 
                                   na.value = "transparent",
                                   name = "PDSI",
                                   limits = c(-5,5),
                                   n.breaks = 9) +
      geom_polygon(data = land %>% st_crop(land_rast) %>% as("Spatial") %>% fortify(), 
                 fill = NA,
                 color = "black",
                 aes(x = long, y = lat, group = group)) +
      
      guides(fill = guide_colorsteps(barwidth = 15, barheight = 0.5)) +
      theme(legend.position = "bottom") +
      
      labs(x = NULL,
           y = NULL) +
      
      coord_equal()

  }

```

The differences mentioned above can better be appreciated in the following figure. The most notable area of divergence is in the Iberian peninsula between the REMO ensemble and vdS. When comparing ERA5 and vdS, eastern Europe/Russia also displays considerable divergence between datasets.

```{r fig_2_2_2, fig.height=4.5}

pmap(ds, function(ds1, ds2){

  str_glue("{models_s[ds1]}_vs_{models_s[ds2]}") -> var_name
  
  all_pdsi_percs_seas %>%
    slice(q, 2) %>%
    split("model") %>% 
    select(models_s) %>% 
    rename("var1" = ds1,
           "var2" = ds2) %>% 
    mutate(dif = var1 - var2) %>% 
    select(dif) %>% 
    setNames(var_name)
    
}) -> tb_comp

tb_comp %>% 
  do.call(c, .) %>% 
  merge(name = "comp") -> mann_comparisons

mann_comparisons %>% 
  st_get_dimension_values("comp") %>% 
  str_replace("_vs_", " - ") %>% 
  {st_set_dimensions(mann_comparisons, "comp", values = .)} -> mann_comparisons

mann_comparisons %>%
  mutate(X = case_when(X > 5 ~ 5,
                       X < -5 ~ -5,
                       TRUE ~ X)) -> mann_comparisons

ggplot() +
  geom_stars(data = mann_comparisons) +
  facet_grid(comp %>% factor(levels = comp_levels) ~ season) +
  scale_fill_binned_diverging(palette = "Vik",
                              rev = T,
                              na.value = "transparent",
                              limits = c(-5,5),
                              n.breaks = 11,
                              name = "difference\n(PDSI)") +
  
  theme(strip.text.x = element_text(size = 7.5)) +
  geom_polygon(data = land %>% st_crop(land_rast) %>% as("Spatial") %>% fortify(), 
               fill = NA,
               color = "black",
               aes(x = long, y = lat, group = group)) +
  
  guides(fill = guide_colorsteps(barwidth = 15, barheight = 0.5)) +
  theme(legend.position = "bottom") +
  
  labs(x = NULL,
       y = NULL) +
  
  coord_equal()

```









